Handling Recursion in Generic Programming Using Closed Type Families

### 2018-08-04 lncs-paper.tex
Address small-scale remarks, some wibble

We feel both existing approaches unsatisfactory and make the following contributions.
\begin{itemize}
\item We describe the problem with the current approach of SOP in detail (Section~\ref{sec:sop-problem}).
\item We introduce an idiom that overcomes the problem. The approach avoids both, the use of overlapping instances and changing a generic representation (Section~\ref{sec:handling-recursion}).
% -- 
%\item We provide a convenient interface for generic functions: there is no need to manually declare type class instances for using them.
% -- this doesn't sound like a contribution
\item We evaluate our approach through the development of a larger-scale use case---the generic zipper. The zipper is meant to be easily and flexibly used with families of mutually recursive datatypes (Section~\ref{sec:generic-zipper}).
\item We note, that our approach can contribute to the \textsf{generics-sop}'s one eliminating some boilerplate instance declarations, which necessarily  arise in practice as a consequence of absence of information about recursion points. An example of that, taken from the \textsf{basic-sop}~\cite{basic-sop} package, is discussed in Section~\ref{subsec:gshow}.
\end{itemize}

@@

Consider, for instance, a datatype of binary trees:
\begin{lstlisting}
data Tree a = Leaf a | Node (Tree a) (Tree a)
\end{lstlisting}
This datatype is isomorphic to the following code:
\begin{lstlisting}
type family Code (Tree a) = '[ '[a], '[Tree a, Tree a]]
\end{lstlisting}

@@

We leave out discussion of the \K{SListI} constraint in the \K{Generic} class definition as irrelevant to our work. Although, we do use \K{All} constraint combinator (as in \K{All SListI}) in the following. Therefore, it is worth noting that \K{All} applies a particular constraint (e.g. \K{SListI} above) to each member of a list of types. The usage of constraints as type arguments is allowed due to \textsf{ConstraintKinds} language extension introducing a dedicated kind \K{Constraint}.

@@

We illustrate the problem through a short example. The \textsf{QuickCheck} library~\cite{Claessen2011} for automatic testing of Haskell code defines a helper function \K{subterms} that takes a term and obtains a list of all its immediate subterms that are of the same type as the given term, that is, all the recursive positions in the term structure. We reimplement this function using the SOP view:
\begin{lstlisting}
subterms :: Generic a => a -> [a]
subterms t = subtermsNS $ unSOP $ from t

@@

subtermsNP :: ∀a xs. NP I xs -> [a]
subtermsNP p (I y :* ys)
  | typeEq @a y = witnessEq y : subtermsNP ys
  | otherwise   = subtermsNP ys
subtermsNP _ Nil = []
\end{lstlisting}
The function \K{subterms} translates the term to its representation, unwrapping the sum of products from \K{SOP}, and passes that to the auxiliary function \K{subtermsNS}. The latter merely traverses the sum and, once reaches the product,  passes it further to \K{subtermsNP}.

@@

@@

Closed type families are the Haskell language extension introduced by Eisenberg et al.~\cite{Eisenberg2014}. The main idea of the extension is that the equations for a \emph{closed type family} are disallowed outside its declaration. Under the extension, we can give the following definition of type-level equality:
\begin{lstlisting}
type family Equal a x :: Bool where
  Equal a a = 'True
  Equal a x = 'False
\end{lstlisting}
The equations in a closed type family are matched in a top-to-bottom order. Since the order is fixed, the overlapping equations here cannot be used to define unsound type-level equations.

@@

We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality, we can witness the coercion between equal types by defining a type class:
\begin{lstlisting}
class Proof (eq :: Bool) (a :: *) (b :: *) where
  witnessEq :: b -> Maybe a

instance Proof 'False a b where
  witnessEq _ = Nothing
instance Proof 'True  a a where
  witnessEq   = Just
\end{lstlisting}

For every element in a list of all direct subterms of a term we shall provide a proof object witnessing its type (in)equality to the type of the term. This can be done by means of the \K{All} combinator and partially applied auxiliary type class \K{ProofEq}, which abbreviates the heavy-weighted interface of \K{Proof}:
\begin{lstlisting}
class    Proof (Equal a b) a b => ProofEq a b
instance Proof (Equal a b) a b => ProofEq a b
\end{lstlisting}
Resulting implementation of \K{subtermsNP} resembles our first definition given in the previous section:
\begin{lstlisting}
subtermsNP :: ∀a xs. All (ProofEq a) xs => NP I xs -> [a]
subtermsNP (I (y :: x) :* ys)
  = case witnessEq @(Equal a x) y of
      Just t  -> t : subtermsNP ys
      Nothing -> subtermsNP  ys
subtermsNP _ Nil = []
\end{lstlisting}

As a side note, we make use of \textsf{ScopedTypeVariables} extension in the definition above, as the type of the element being matched does not appear in the function signature, since it may match an empty list.

@@

For managing polymorphic recursion in this datatype, we can write an analogue of the \K{Equal} type family from Section~\ref{sec:handling-recursion}, which ignores type parameters when checking two polymorphic types. Since any datatype with type parameters \K{f a b c ...} has kind \K{* -> (* -> (* -> ...))}, a \K{PolyEq} type family can be defined thus:
\begin{lstlisting}
type family PolyEq (a :: k) (x :: k) :: Bool where
  PolyEq (f a) (g b) = PolyEq f g
  PolyEq a     a     = 'True
  PolyEq a     x     = 'False
\end{lstlisting}

The function \K{gshow} from Section~\ref{subsec:gshow} can be reimplemented by using this type family instead of \K{Equal}. The only piece of its definition must be changed as well in order to recursively invoke the function each time with a new type:
\begin{lstlisting}
instance Show  b => CaseShow 'False a b where
  caseShow'   = show
instance GShow b => CaseShow 'True  a b where
  caseShow' t = "(" ++ gshow t ++ ")"
\end{lstlisting}

Unfortunately, this approach, albeit working well for datatypes defined like \K{PolyRec~a}, becomes unsuitable for nested datatypes, such as one below:
\begin{lstlisting}
data Nested a = Epsilon | Nest a (Nested [a])
\end{lstlisting}
The culprit is the constraint \K{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the constraint, because of the \K{GShow b} instance context above.

@@

The classical zipper described by Huet~\cite{Huet1997} can be generically calculated for regular datatypes~\cite{HiJeLo2004}---a subset of datatypes that can be viewed as a least fixed point of some polynomial expression on types. Yakushev~et~al.~\cite{MuRec2009} generalize the definition of the generic zipper for an arbitrary family of mutually recursive datatypes. All mentioned solutions require a datatype to be expressed using forms of a fixed-point operator, since the zipper operates on recursion points.

@@

One last missing piece for managing focuses is the class \K{ProofIn}. It provides a proof of membership of a focus type to a family. Again, this generalizes the proof of type equality from Section~\ref{sec:handling-recursion}. The definitions of \K{ProofIn} and an auxiliary class \K{ProofFocus} are given in Figure~\ref{fig:proof-in}. 
%The constraints on the second \K{ProofFam} instance, which repeat the same ones from the \K{Family} constructor, constitute the proof.
% this is not clear and, I think, can be omitted

@@

type family (.*) (x :: *) (ys :: [[*]]) :: [[*]] where
  x .* (ys ': yss) = (x ': ys) ': (x .* yss)
  x .* '[]         = '[]
infixr 7 .*

@@

type family DiffProd (fam :: [*]) (xs :: [*]) :: [[*]] where
  DiffProd fam '[]       = '[]
  DiffProd fam '[x]      = If (InFam x fam) '[ '[Hole]] '[]
  DiffProd fam '[End, x] = If (InFam x fam) '[ '[]]     '[]
  DiffProd fam (x ': xs)
    = Hole .* xs .** DiffProd fam '[End, x] .++
              x  .*  DiffProd fam xs
\end{lstlisting}}
\end{CenteredBox}
\caption{Differentiation of a product of types.}
\label{fig:diff-prod}
\end{figure}

@@

type family ToContext (n :: ConsN) (fam :: [*])
                      (code :: [[*]]) :: [[*]] where
  ToContext n fam '[] = '[]
  ToContext n fam (xs ': xss)
    = Proxy n .* DiffProd fam xs .++ ToContext ('N n) fam xss
\end{lstlisting}
The type family \K{ToContext} derives the type of the context of a datatype performing differentiation of a sum on its code. Since each product from the code matches a sum of multiple products in the context, it is helpful for each product of the context representation, to keep the index of its matching constructor of the datatype. We store the index in the datatype \K{ConsN} adding that to the head of each product but wrapped by \K{Proxy} because we use \K{ConsN} promoted to a kind in the type family, while the products contain types of kind \K{*}. The constructors \K{F} and \K{N} denote ``first'' and ``next'', respectively, and the special constructor \K{None} will be used further to indicate failure of matching indices.

@@

Once we have constructed the product, we have to build the sum representing the choice of that product. If we have the index of the chosen constructor for the datatype, the one that we find for its context can be computed as follows:
\begin{lstlisting}
type family CtxConsN (xss :: [[*]]) (n :: ConsN) :: ConsN where
  CtxConsN '[]                       n = 'None
  CtxConsN ((Proxy n  ': xs) ': xss) n = 'F
  CtxConsN ((Proxy n' ': xs) ': xss) n = 'N (CtxConsN xss n)
\end{lstlisting}
The list \K{xss} here is expected to be the context code, which this traverses until the first product storing the constructor index equal to the given one, i.~e., the context's first constructor matching with the chosen constructor of the datatype.

@@

We now can witness the choice using the proof of type equality we have defined in Section~\ref{sec:handling-recursion}. In the following definition, \K{f} is supposed to be instantiated to \K{Inj (NP I) xss} where \K{xss} reflects the code of the context\footnote{This might be simplified by using a singleton type for \K{ConsN} instead of defining a type class function. However, when using a singleton, the recursive definition of \K{toFirstCtx} we give further requires a constraint on its function signature that leads to a nonterminating computation.}:
\begin{lstlisting}
class    Proof (Equal a b) (f a) (f b) => ProofF f a b
instance Proof (Equal a b) (f a) (f b) => ProofF f a b

@@

instance ConsNInj n xs => ConsNInj ('N n) xs where
  consNInj (_ :* xss) = consNInj @n xss
  consNInj Nil        = impossible
instance ConsNInj 'F xs where
  consNInj ((xs :: f xs) :* _)
    = fromMaybe impossible $ witnessEq @(Equal xs ys) xs
  consNInj Nil = impossible
instance ConsNInj 'None xs where
  consNInj _   = impossible

@@

Several works show how to define the Zipper~\cite{Huet1997} generically for regular~\cite{HiJeLo2004,McBride2001} and mutually recursive~\cite{MuRec2009} types using fixed-point generic views. Adams~\cite{Adams2010} defines a generic zipper for heterogeneous types.

>>>>

The problem of using overlapping instances was first addressed by Kiselyov~et~al.~\cite{Kiselyov2004}. Their technique for avoiding overlap relies on a Haskell 98 extension for functional dependencies. The solution proposes two variants of defining a type-level equality predicate, a type class, and then systematically localizes overlap by circumventing it with that predicate. The first version of type equality maps types to unique type representations and compares them. Its later implementation with type families~\cite{Kiselyov2012} fully eliminates \textsf{OverlappingInstances}. Although, each type needs a representation instance to be derived---by means of Template Haskell or GHC. The most generic solution for type equality, the second version, again makes use of overlapping instances, however.

Closed type families, today's Haskell extension, has been proposed primarily to obviate the need for overlapping instances. But it does not seem to be well-known that they have useful application in the generic programming area.

We follow up the overlap-avoiding idea from the existing approach and make the following contributions.
\begin{itemize}
\item We describe the problem with the current approach of SOP in detail (Section~\ref{sec:sop-problem}).
\item We introduce an idiom that overcomes the problem. The approach avoids both, the use of overlapping instances and changing a generic representation (Section~\ref{sec:handling-recursion}).
% -- 
%\item We provide a convenient interface for generic functions: there is no need to manually declare type class instances for using them.
% -- this doesn't sound like a contribution
\item We evaluate our approach through the development of a larger-scale use case---the generic zipper. The zipper is meant to be easily and flexibly used with families of mutually recursive datatypes (Section~\ref{sec:generic-zipper}).
\item We note, that our approach can contribute to the \textsf{generics-sop}'s one eliminating some boilerplate instance declarations, which necessarily  arise in practice as a consequence of absence of information about recursion points. An example of that, taken from the \textsf{basic-sop}~\cite{basic-sop} package, is discussed in Section~\ref{subsec:gshow}.
\end{itemize}

@@

Consider, for instance, a datatype of binary trees:
\begin{lstlisting}
data Tree a = Leaf a | Node (Tree a) (Tree a)
\end{lstlisting}
This datatype is isomorphic to the following code:
\begin{lstlisting}
type instance Code (Tree a) = '[ '[a], '[Tree a, Tree a]]
\end{lstlisting}

@@

We leave out discussion of the \K{SListI} constraint in the \K{Generic} class definition as irrelevant to our work. Although, we do use \K{All} constraint combinator (as in \K{All SListI}) in the following. Therefore, it is worth noting that \K{All} applies a particular constraint (e.g. \K{SListI} above) to each member of a list of types. The usage of constraints as type arguments is allowed due to the \textsf{ConstraintKinds} language extension introducing a dedicated kind \K{Constraint}.

@@

We illustrate the problem through a short example. The \textsf{QuickCheck} library~\cite{Claessen2011} for automatic testing of Haskell code defines a helper function \K{subterms} that takes a term and obtains a list of all its immediate subterms that are of the same type as the given term, that is, all the recursive positions in the term structure. In the following, we reimplement this function using the SOP view. But first we give a sketch of solution to introduce the idea.

\subsubsection{Non-implementation}

The SOP-based \K{subterms} definition is outlined below:
\begin{lstlisting}
subterms :: Generic a => a -> [a]
subterms t = subtermsNS $ unSOP $ from t

@@

subtermsNP :: ∀a xs. NP I xs -> [a]
subtermsNP p (I y :* ys)
  | typeOf @a y = castEq y : subtermsNP ys
  | otherwise   = subtermsNP ys
subtermsNP _ Nil = []
\end{lstlisting}
The function \K{subterms} translates the term to its representation, unwrapping the sum of products from \K{SOP}, and passes that to the auxiliary function \K{subtermsNS}. The latter merely traverses the sum and, once reaches the product,  passes it further to \K{subtermsNP}.

@@

\subsubsection{Overlap-based implementation}

@@

Closed type families are the Haskell language extension introduced by Eisenberg et al.~\cite{Eisenberg2014}. The main idea of the extension is that the equations for a \emph{closed type family} are disallowed outside its declaration. Under the extension, we can give the following definition of type-level equality:
\begin{lstlisting}
type family Equal a x :: Bool where
  Equal a a = 'True
  Equal _ _ = 'False
\end{lstlisting}
The equations in a closed type family are matched in a top-to-bottom order. Since the order is fixed, the overlapping equations here cannot be used to define unsound type-level equations.

@@

We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality predicate, we can decide if \K{a == b} by defining a type class:
\begin{lstlisting}
class DecideEq (eq :: Bool) (a :: *) (b :: *) where
  decideEq :: Maybe (b :==: a)

instance a == b => DecideEq True  a b where
  decideEq = Just Refl
instance          DecideEq False a b where
  decideEq = Nothing
\end{lstlisting}
If the types \K{a} and \K{b} are the same, the \K{:==:} type from \K{Data.Type.Equality} witnesses the equality.

For every element in a list of all direct subterms of a term we shall provide a proof object witnessing its type (in)equality to the type of the term. This can be done by means of the \K{All} combinator and partially applied auxiliary type class \K{ProofEq}, which abbreviates the heavy-weighted interface of \K{DecideEq}:
\begin{lstlisting}
class    DecideEq (Equal a b) a b => ProofEq a b
instance DecideEq (Equal a b) a b => ProofEq a b
\end{lstlisting}
The equality proof can then be employed to provide a type-safe cast between two equal types:
\begin{lstlisting}
castEq :: ∀a b. ProofEq a b => b -> Maybe a
castEq t = (\d -> castWith d t) <$> decideEq @(Equal a b)
\end{lstlisting}
Resulting implementation of \K{subtermsNP} resembles our first definition given in the previous section:
\begin{lstlisting}
subtermsNP :: ∀a xs. All (ProofEq a) xs => NP I xs -> [a]
subtermsNP (I (y :: x) :* ys)
  = case castEq y of
      Just t  -> t : subtermsNP ys
      Nothing -> subtermsNP  ys
subtermsNP _ Nil = []
\end{lstlisting}

As a side note, we make use of the \textsf{ScopedTypeVariables} extension in the definition above, as the type of the element being matched does not appear in the function signature, since it may match an empty list.

@@

For managing polymorphic recursion in this datatype, we can write an analogue of the \K{Equal} type family from Section~\ref{sec:handling-recursion}, which ignores type parameters when checking two polymorphic types. Since any datatype with type parameters \K{f a b c ...} has kind \K{* -> (* -> (* -> ...))}, a \K{PolyEq} type family can be defined thus:
\begin{lstlisting}
type family PolyEq (a :: k) (x :: k) :: Bool where
  PolyEq (f a) (g b) = PolyEq f g
  PolyEq a     a     = 'True
  PolyEq _     _     = 'False
\end{lstlisting}

The function \K{gshow} from Section~\ref{subsec:gshow} can be reimplemented by using this type family instead of \K{Equal}. The only piece of its definition must be changed as well in order to recursively invoke the function each time with a new type (with the proper change in the \K{CaseRecShow} definition):
\begin{lstlisting}
instance Show  a => CaseShow 'False a where
  caseShow'   = show
instance GShow a => CaseShow 'True  a where
  caseShow' t = "(" ++ gshow t ++ ")"
\end{lstlisting}

Unfortunately, this approach, albeit working well for datatypes defined like \K{PolyRec a}, becomes unsuitable for nested datatypes, such as one below:
\begin{lstlisting}
data Nested a = Epsilon | Nest a (Nested [a])
\end{lstlisting}
The culprit is the constraint \K{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the constraint, because of the \K{GShow b} instance context above.

@@

The classical zipper described by Huet~\cite{Huet1997} can be generically calculated for regular datatypes~\cite{HiJeLo2004}---all datatypes expressible as polynomial expressions on types. Yakushev~et~al.~\cite{MuRec2009} generalize the definition of the generic zipper for an arbitrary family of mutually recursive datatypes. All mentioned solutions require a datatype to be expressed using forms of a fixed-point operator, since the zipper operates on recursion points.

@@

One last missing piece for managing focuses is the class \K{ProofIn}. It provides a proof of membership of a focus type to a family. Again, this generalizes the proof of type equality from Section~\ref{sec:handling-recursion}. The definitions of \K{ProofIn} and an auxiliary class \K{ProofFocus} are given in Figure~\ref{fig:proof-in}. In the setting of families, where there are no analogues of \K{:==:}, we can define only a weak form of proof, which is more flexible.
%The constraints on the second \K{ProofFam} instance, which repeat the same ones from the \K{Family} constructor, constitute the proof.
% this is not clear and, I think, can be omitted

@@

type family (.*) (x :: *) (ys :: [[*]]) :: [[*]] where
  x .* (ys ': yss) = (x ': ys) ': (x .* yss)
  _ .* '[]         = '[]
infixr 7 .*

@@

type family DiffProd (fam :: [*]) (xs :: [*]) :: [[*]] where
  DiffProd _   '[]       = '[]
  DiffProd fam '[x]      = If (InFam x fam) '[ '[Hole]] '[]
  DiffProd fam '[End, x] = If (InFam x fam) '[ '[]]     '[]
  DiffProd fam (x ': xs)
    = Hole .* xs .** DiffProd fam '[End, x] .++
              x  .*  DiffProd fam xs
\end{lstlisting}}
\end{CenteredBox}
\caption{Differentiation of a product of types.}
\label{fig:diff-prod}
\end{figure}

@@

type family ToContext (n :: ConsN) (fam :: [*])
                      (code :: [[*]]) :: [[*]] where
  ToContext _ _ '[] = '[]
  ToContext n fam (xs ': xss)
    = Proxy n .* DiffProd fam xs .++ ToContext ('N n) fam xss
\end{lstlisting}
The type family \K{ToContext} derives the type of the context of a datatype performing differentiation of a sum on its code. Since each product from the code matches a sum of multiple products in the context, it is helpful for each product of the context representation, to keep the index of its matching constructor of the datatype. We store the index in the datatype \K{ConsN} adding that to the head of each product but wrapped by \K{Proxy} because we use \K{ConsN} promoted to a kind in the type family, while the products contain types of kind \K{*}. The constructors \K{F} and \K{N} denote ``first'' and ``next'', respectively, and the special constructor \K{None} will be used further to indicate failure of matching indices.

@@

Once we have constructed the product, we have to build the sum representing the choice of that product. If we have the index of the chosen constructor for the datatype, the one that we find for its context can be computed as follows:
\begin{lstlisting}
type family CtxConsN (xss :: [[*]]) (n :: ConsN) :: ConsN where
  CtxConsN '[]                      _ = 'None
  CtxConsN ((Proxy n  ': _) ': _  ) n = 'F
  CtxConsN ((Proxy n' ': _) ': xss) n = 'N (CtxConsN xss n)
\end{lstlisting}
The list \K{xss} here is expected to be the context code, which this traverses until the first product storing the constructor index equal to the given one, i.~e., the context's first constructor matching with the chosen constructor of the datatype.

@@

We can witness the choice using a variant of the type-safe cast we have defined in Section~\ref{sec:handling-recursion}, which takes the result of the equality predicate:
\begin{lstlisting}
typeCastEq :: ∀eq a b. DecideEq eq a b => b -> Maybe a
\end{lstlisting}
In the following definition, \K{f} is supposed to be instantiated to \K{Inj (NP I) xss} where \K{xss} reflects the code of the context\footnote{This might be simplified by using a singleton type for \K{ConsN} instead of defining a type class function. However, when using a singleton, the recursive definition of \K{toFirstCtx} we give further requires a constraint on its function signature that leads to a nonterminating computation.}:
\begin{lstlisting}
class    DecideEq (Equal a b) (f a) (f b) => ProofF f a b
instance DecideEq (Equal a b) (f a) (f b) => ProofF f a b

@@

instance ConsNInj n xs => ConsNInj ('N n) xs where
  consNInj (_ :* xss) = consNInj @n xss
  consNInj Nil        = impossible
instance ConsNInj 'F xs where
  consNInj ((xs :: f xs) :* _)
    = fromMaybe impossible $ typeCastEq @(Equal xs ys) xs
  consNInj Nil = impossible
instance ConsNInj 'None xs where
  consNInj _ = impossible

@@

Several works show how to define the Zipper~\cite{Huet1997} generically for regular~\cite{HiJeLo2004,McBride2001} and mutually recursive~\cite{MuRec2009} types using fixed-point generic views. Adams~\cite{Adams2010} defines a generic zipper for heterogeneous types: a different kind of zipper that can traverse knots of various types, so it does not use the recursive structure.
